---
title: "Nonlinear Optimization Ex.1"
author: "Ariel Vishne 204149371"
date: "22 3 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

## Ex. 1
As seen in class (week #2), we require solving the equation:
$$
    x - 2sin(x) = 0
$$
We saw that the segment $[\frac{\pi}{2}, 2]$ has some positive root. We will find it with bisection and Newton methods.

### Ex. 1-i BISECTION
```{r ex1-bisection}

f <- function(x){
  return(x - 2*sin(x))
}
precision <- 10^-6


a <- (pi / 2)
b <- 2
i <- 0
x <- ((a+b)/2)

while(abs(a-b) > precision){
  i <- i+1
  print(paste("iteration number", i, "current segment [", a,",", b, "]", "Current precision", round(abs(a-b),7)))
  if(f(x)*f(a)<0){
    b <- x
  }
  else if(f(x)*f(b) <0){
    a <- x
  }
  x <- ((a+b)/2)
}

```
### Ex. 1-ii NEWTON
```{r ex1-newton}
x <- 2
i <- 0

while(f(x) > precision){
  i <- i+1
  print(paste("iteration number", i, "precision reached:", round(abs(f(x)),7), "value reached:", round(x,7)))


  x <- x - ((x-2*sin(x))/(1-2*cos(x)))
}

```
### Ex. 1-iii Comparison
We can see that the Newton method converges sqared, while bisection converges linear.

## Ex. 2
The desired quantity $\frac{1}{\sqrt2}$ is a solution (a root) to the equation
$$
    (\sqrt2x - 1)^2 = 0
$$
This equation has no division. We therefore will use the Newton method beginning at ${X_0 = 1/2}$.

If we take the derivative we have:

$$
((\sqrt2x - 1)^2)'=2\sqrt2(\sqrt2x-1)=4x-2\sqrt2
$$
If we solve the recursive equation and we multiply by $2\sqrt2$ we reach:
$$
X_{n+1} = X_{n} - \frac{2\sqrt2f(x)}{f'(x)}=X_n- \frac{2\sqrt2((\sqrt2x-1)^2)}{2\sqrt2(\sqrt2x-1)}=X_{n}-[\sqrt2x-1]=X_{n}-\sqrt2x+1
$$
Thus we have reached an equation for which $\frac{1}{\sqrt2}$ is a root, and which has no division.
Beginning at $x_0=0.5$ we get:

```{r ex-2}
f <- function(x){
  return((sqrt(2)*x - 1)^2)
}
precision <- 10^-6
x <- 1/2
i <- 0

while(abs(f(x)) > precision){
  i <- i+1
  print(paste("iteration number", i, "precision reached:", round(abs(f(x)),7), "value reached:", round(x,7)))


  x <- (x - (sqrt(2)*x) + 1)
}

```
We therefore conclude that the value $0.7071078$ is an approximation of $\frac{1}{\sqrt2}$ by six significant digits.

## Ex. 3

We have
$$
    f(x) = e^x - x - 1
$$
if we look at the derivative we have
$$
f'(x) = (e^x - x - 1)' = e^x - 1
$$

Since $e^x-1$ has only one root where $e^{x} - 1 =0\Rightarrow e^{x}=1\Rightarrow x = 0$ we only have one maxima/minima point. Since $f''(x) = e^{x} > 0$ for all $x$ then it is a minimum point. We conclude that $x = 0$ is the only root.

### Ex. 3-i Newton
To compare the methods we will begin at $X_0 = 0.125$ and assume precision required of $10^{-4}$
For the regular Newton method we calculate:
$$
    X_{n+1} = X_{n} - \frac{f(X_n)}{f'(X_n)} = X_n -\frac{e^{X_n} - X_n - 1}{e^{X_n}} = X_n - 1 + \frac{X_n + 1}{e^{X_n}}
$$


```{r ex3-newton}

f <- function(x){
  return((exp(1)^x) - x - 1)
}
precision <- 10^-4
x <- 1/8
i <- 0

while(abs(f(x)) > precision){
  i <- i+1
  print(paste("iteration number", i, "precision reached:", round(abs(f(x)),7), "value reached:", round(x,7)))

  x <- (x - 1) + ((x+1) / (exp(1)^x))
}


```
We required 127 iterations to get to the required precision.

### Ex 3-ii
If however we multiply the numerator by 2 we obtain:
$$
    X_{n+1} = X_{n} - \frac{2f(X_n)}{f'(X_n)} = X_n -\frac{2e^{X_n} - 2X_n - 2}{e^{X_n}} = X_n - 2 + \frac{2X_n + 2}{e^{X_n}}
$$

```{r ex3-newton-modified}

f <- function(x){
  return((exp(1)^x) - x - 1)
}
precision <- 10^-4
x <- 1/8
i <- 0

while(abs(f(x)) > precision){
  i <- i+1
  print(paste("iteration number", i, "precision reached:", round(abs(f(x)),7), "value reached:", round(x,7)))

  x <- (x - 2) + (((2*x)+2) / (exp(1)^x))
}


```
We require only 63 iterations, exactly half of those required in the previous method

### Ex. 3-iii
The reason that the second method works better is because $f(x)$ has the first derivative also obtaining the value of 0 at point 0 ($f'(0)=0$), it is better to multiply by the order of the first derivative that does not recieve the value of 0. Because $f''(0)=e^0=1\ne0$ then multiplying the numerator by 2 will achieve faster convergence.

## Ex. 4
We have
$$
f(x) = arctan(x) \Rightarrow f'(x) = \frac{1}{1+x^2}
$$
Therefore using the Newton method we obtain:
$$
X_{n+1} = X_n - \frac{f(X_n)}{f'(X_n)}= X_n - \frac{arctan(X_n)}{\frac{1}{1+X_n^2}} = X_n- (1+X_n^2)arctan(X_n) 
$$
For convergence to 0 we require $|X_{n+1}|<|X_{n}|$ and thus we require that
$$
X_n- (1+X_n^2)arctan(X_n) > -X_n\Longrightarrow (1+X_n^2)arctan(X_n) > 2X_n \Longrightarrow |arctan(X_n)| > |\frac{2X_n}{1+X_n^2}|
$$
We can thus see that for the domain of
$$
-1.39<X_n<1.39
$$
Where $X\ne0$ there will be a convergence. For example, we check for $x=1.35$ and obtain the following results:
```{r ex4}
f <- function(x){
  return(atan(x))
}
precision <- 10^-6
x <- 1.35
i <- 0

while(abs(f(x)) > precision){
  i <- i+1
  print(paste("iteration number", i, "precision reached:", round(abs(f(x)),7), "value reached:", round(x,7)))
  

  x <- x - ((x^2)*f(x)) - f(x)
}

t <- seq(-5,5,0.01)
plot(t, abs(atan(t)), col = "red", xlab = "", ylab = "")
points(t, abs(((2*t)/(1+(t^2)))), col = "blue")
abline(v = c(-1.39,1.39), col = c("green", "green"))
legend("topleft", c("|atan|", "|2x/1+x^2|", "convergence zone"), fill = c("red", "blue", "green"))
```